---
title: "RQ3test"
output: html_document
date: "2026-09-03"
---
```{r}
library(readr)
library(dplyr)
library(stringr)
library(tidyverse)
library(here)
```


```{r}
df1 <- read.csv(here("data", "AHSnhs11bsp.csv"), stringsAsFactors = TRUE)
df2 <- read.csv(here("data", "AHSnhs11bbi.csv"), stringsAsFactors = TRUE)
                
df <- df1 %>%
  filter(!(ABSLID == "NHS11B0009794" & ABSPID == 6)) %>%
  select(ABSLID, ABSPID, SEX, ALCSTR01, AL2K7DAY) %>%
  left_join(
    df2 %>%
    filter(!(ABSLID == "NHS11B0009794" & ABSPID == 6)) %>%
    select(ABSLID, ABSPID, EGFRKDSA, ACRNTR, ACRIPA, EGFRIMP, EGFRRESB),
    by = c("ABSLID", "ABSPID")
  )
```

```{r}
# Duplicate id
df1 %>%
  count(ABSLID, ABSPID) %>%
  filter(n > 1)
```

```{r}
# Check the number of observations in each category
table(df$FERRCRPB, useNA = "ifany")
```


```{r}
df %>%
  count(EGFRKDSA, SEX)
```


```{r}
# Some categories were combined to simplify EDA
ckd_data <- df %>%
  mutate(
    
    SEX = case_when(
      SEX == 1 ~ "Male",
      SEX == 2 ~ "Female",
    ),
    CKD = case_when(
      EGFRKDSA == 1 ~ "No CKD",
      EGFRKDSA %in% 2:6 ~ "CKD",
      EGFRKDSA %in% c(0, 8) ~ NA_character_
    ),
    ACR_status = case_when(
      ACRNTR == 1 ~ "No albuminuria",
      ACRNTR %in% c(2, 3) ~ "Albuminuria",
      ACRNTR %in% c(0, 8) ~ NA_character_
    ),
    ACR_ratio_status = case_when(
      ACRIPA == 1 ~ "Normoalbuminuria",
      ACRIPA == 2 ~ "Microalbuminuria",
      ACRIPA == 3 ~ "Macroalbuminuria",
      ACRIPA %in% c(0, 8) ~ NA_character_
    ),
    eGFR_status = case_when(
      EGFRIMP == 1 ~ "Normal",
      EGFRIMP == 2 ~ "Impaired",
      EGFRIMP %in% c(0, 8) ~ NA_character_
    ),
    Alcohol_risk = case_when(
    ALCSTR01 %in% c(1, 2) ~ "High risk",
    ALCSTR01 %in% c(3, 4) ~ "Risky",
    ALCSTR01 %in% c(5, 6) ~ "Low risk",
    ALCSTR01 == 8 ~ "No alcohol (past year)",
    ALCSTR01 == 9 ~ "Never alcohol",
    ALCSTR01 %in% c(0, 7) ~ NA_character_
  ),
    Alcohol_risk_7day = case_when(
    AL2K7DAY == 1 ~ "Low risk",
    AL2K7DAY == 2 ~ "Medium risk",
    AL2K7DAY == 3 ~ "High risk",
    AL2K7DAY == 4 ~ "No alcohol (1 week–<12 months)",
    AL2K7DAY == 5 ~ "No alcohol (≥12 months)",
    AL2K7DAY == 6 ~ "Never alcohol",
    AL2K7DAY %in% c(0, 8, 9) ~ NA_character_
  )
  )
```



```{r}
# Calculate the proportion within each sex
ckd_prop <- ckd_data %>%
  filter(!is.na(SEX), !is.na(CKD)) %>%
  group_by(SEX, CKD) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(SEX) %>%
  mutate(proportion = n / sum(n))
```


```{r}
ggplot(ckd_prop, aes(x = factor(CKD),
                    y = proportion,
                    fill = factor(SEX))) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    y = "Proportion",
    fill = "Sex"
  ) +
  theme_minimal()
ggsave(here("figures", "sex_CKD.png"))
```


```{r}
alcohol_type <- read.csv(here("data", "AHSnhs11b14.csv"), stringsAsFactors = TRUE) %>%
  select(
    ABSLID,
    ABSPID,
    ABSBID,
    ABSTID,
    ALCTYPT,
    TOTPALCR
  )
```

```{r}
alcohol_quantity <- df1 %>%
  select(
    ABSLID,
    ABSPID,
    ALINWKSD
  ) %>%
  
  mutate(
    alcohol_quantity = case_when(
      ALINWKSD == 97 ~ 0,          # Did not drink last week
      ALINWKSD >= 0 & ALINWKSD <= 96 ~ ALINWKSD,  # Actual alcohol intake
      ALINWKSD %in% c(98, 99) ~ NA_real_  # Not stated/unknown or not applicable
    )
  )
```

```{r}
alcohol_type %>%
  count(ALCTYPT) #Counts of type of alcohol consumed, 0 means did not drink last week 
```

```{r}
alcohol_type <- alcohol_type %>%
  mutate(
    alcohol_type = case_when(
      ALCTYPT == 1 ~ "Low_alcohol_beer",
      ALCTYPT == 2 ~ "Mid_strength_beer",
      ALCTYPT == 3 ~ "Full_strength_beer",
      ALCTYPT == 4 ~ "White_wine",
      ALCTYPT == 5 ~ "Red_wine",
      ALCTYPT == 6 ~ "Low_alcohol_wine",
      ALCTYPT == 7 ~ "Champagne_sparkling_wine",
      ALCTYPT == 8 ~ "Spirits_RTD",
      ALCTYPT == 9 ~ "Liqueurs",
      ALCTYPT == 10 ~ "Spirits",
      ALCTYPT == 11 ~ "Fortified_wine",
      ALCTYPT == 12 ~ "Cider",
      ALCTYPT == 13 ~ "Cocktails",
      ALCTYPT == 14 ~ "Other",
      ALCTYPT == 0 ~ NA_character_
    )
  )
```

```{r}
# Keep one record for each participant and alcohol type
alcohol_type_person <- alcohol_type %>%
  filter(!is.na(alcohol_type)) %>%
  distinct(ABSLID, ABSPID, alcohol_type) %>%
  mutate(value = 1) %>%
  pivot_wider(
    names_from = alcohol_type,
    values_from = value,
    values_fill = 0
  )
```

```{r}
# Combine alcohol type and alcohol quantity at the participant level
alcohol_person <- alcohol_quantity %>%
  left_join(
    alcohol_type_person,
    by = c("ABSLID", "ABSPID")
  )
```

```{r}
# Add alcohol information to the main participant-level dataset
df_final <- ckd_data %>%
  left_join(
    alcohol_person,
    by = c("ABSLID", "ABSPID")
  )

#Note Alcohol_person in df_final only includes people who drank in the last week
```

```{r}
clean_data <- na.omit(df_final)

#Seems like 4 observations were lost after alcohol variables were added 

prop.table(table(clean_data$ACRNTR)) #Albuminuria status
table(clean_data$ACRNTR)

table(clean_data$ACRIPA) #Albumin Creatinine ratio 
prop.table(table(clean_data$ACRIPA))

table(clean_data$EGFRIMP) #eGFR status
prop.table(table(clean_data$EGFRIMP))

table(clean_data$EGFRKDSA) #ckd status
prop.table(table(clean_data$EGFRKDSA))

table(clean_data$CKD)
prop.table(table(clean_data$CKD))


prop.table(table(clean_data$SMKSTAT))
prop.table(table(clean_data$SMKAGEDB))
prop.table(table(clean_data$ALCTYPT))
prop.table(table(clean_data$TOTPALCR))
prop.table(table(clean_data$FOLATREB))
prop.table(table(clean_data$B12RESB))
prop.table(table(clean_data$FERRCRPB))
prop.table(table(clean_data$VDLCMRGT))

#New hack, can just view the clean_data and there's a summary page on proportions
nrow(clean_data)
```

```{r}
nrow(clean_data)
```


```{r}
alcohol_type_prop <- df_final %>%
  filter(!is.na(SEX), !is.na(CKD)) %>%
  pivot_longer(
    cols = c(
      Low_alcohol_beer,
      Mid_strength_beer,
      Full_strength_beer,
      White_wine,
      Red_wine,
      Low_alcohol_wine,
      Champagne_sparkling_wine,
      Spirits_RTD,
      Liqueurs,
      Spirits,
      Fortified_wine,
      Cider,
      Cocktails,
      Other,
    ),
    names_to = "alcohol_type",
    values_to = "consumed"
  ) %>%
  filter(!is.na(consumed), consumed == 1) %>%
  group_by(SEX, alcohol_type, CKD) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(SEX, alcohol_type) %>%
  mutate(proportion = n / sum(n))
```

```{r}

ggplot(
  alcohol_type_prop %>% filter(CKD == "CKD"),
  aes(
    x = alcohol_type,
    y = proportion,
    fill = factor(SEX)
  )
) +
  geom_col(position = "dodge") +
  
  scale_x_discrete(labels = c("Champagne", "Cider", "Cocktails", "Fortified Wine", "Full strength Beer", "Liquers", "Low alcoholic beer", "Low alcoholic wine", "Mid strength Beer","Red Wine","Ready to drink spirits/liquers","Spirits","White Wine" )) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Alcohol type",
    y = "Proportion with CKD",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggsave(here("figures", "alcohol_type_CKD.png"))

```

```{r}
quantity_prop <- df_final %>%
  filter(
    !is.na(SEX),
    !is.na(CKD),
    !is.na(alcohol_quantity)
  ) %>%
  mutate(
    alcohol_group = cut(
      alcohol_quantity,
      breaks = c(-Inf, 0, 1, 2, 4, Inf),
      labels = c("0", "0–1", "1–2", "2–4", ">4")
    )
  ) %>%
  group_by(SEX, alcohol_group) %>%
  summarise(
    proportion_CKD = mean(CKD == "CKD"),
    n = n(),
    .groups = "drop"
  )
```


```{r}
ggplot(
  quantity_prop,
  aes(
    x = alcohol_group,
    y = proportion_CKD,
    fill = SEX
  )
) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Average daily alcohol intake (standard drinks)",
    y = "Proportion with CKD",
    fill = "Sex"
  ) +
  theme_minimal()
ggsave(here("figures", "alcohol_quantity_CKD.png"))
```

```{r}
quantity_prop
```

```{r}
smoker_status_prop <- df_final %>%
  filter(
    !is.na(SEX),
    !is.na(CKD),
    !is.na(smoker_status),
    !is.na(daily_smoke),
  ) %>%
  group_by(SEX, daily_smoke) %>%
  summarise(
    proportion_CKD = mean(CKD == "CKD"),
    n = n(),
    .groups = "drop"
  )
```

```{r}
ggplot(
  smoker_status_prop,
  aes(
    x = daily_smoke,
    y = proportion_CKD,
    fill = SEX
  )
) +
  geom_col(position = "dodge") +
  scale_x_discrete(labels = c("<10 years", "10-29 years", "30 years and over", "Never smoked daily")) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Daily Smoking Duration",
    y = "Proportion with CKD",
    fill = "Sex"
  ) +
  theme_minimal()
ggsave(here("figures", "smoking_daily_CKD.png"))
```

```{r}
smoker_prop <- df_final %>%
  filter(
    !is.na(SEX),
    !is.na(CKD),
    !is.na(smoker_status),
  ) %>%
  group_by(SEX, smoker_status) %>%
  summarise(
    proportion_CKD = mean(CKD == "CKD"),
    n = n(),
    .groups = "drop"
  )
```


```{r}
ggplot(
  smoker_prop,
  aes(
    x = smoker_status,
    y = proportion_CKD,
    fill = SEX
  )
) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Smoking status",
    y = "Proportion with CKD",
    fill = "Sex"
  ) +
  theme_minimal()
ggsave(here("figures", "smoking_CKD.png"))
```


```{r}
micro_prop <- df_final %>%
  filter(
    !is.na(SEX),
    !is.na(CKD)
  ) %>%
  select(SEX, CKD, Folate, Vitamin_B12, Ferritin, Vitamin_D) %>%
  pivot_longer(
    cols = c(Folate, Vitamin_B12, Ferritin, Vitamin_D),
    names_to = "micronutrient",
    values_to = "category"
  ) %>%
  filter(!is.na(category)) %>%
  group_by(micronutrient, SEX, category) %>%
  summarise(
    proportion_CKD = mean(CKD == "CKD"),
    n = n(),
    .groups = "drop"
  )
```



```{r}
ggplot(
  micro_prop,
  aes(
    x = category,
    y = proportion_CKD,
    fill = SEX
  )
) +
  geom_col(position = "dodge") +
  facet_wrap(~ micronutrient, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = NULL,
    y = "Proportion with CKD",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
ggsave(here("figures", "micronutrient_CKD.png"))
```

```{r}
# Alcohol risk
risk_prop <- df_final %>%
  filter(
    !is.na(SEX),
    !is.na(CKD),
    !is.na(Alcohol_risk)
  ) %>%
  group_by(SEX, Alcohol_risk) %>%
  summarise(
    proportion_CKD = mean(CKD == "CKD"),
    n = n(),
    .groups = "drop"
  )

ggplot(
  risk_prop,
  aes(
    x = Alcohol_risk,
    y = proportion_CKD,
    fill = SEX
  )
) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Alcohol risk category",
    y = "Proportion with CKD",
    fill = "Sex"
  ) +
  theme_minimal()
```

```{r}
# Alcohol risk level 7 day average(2001)
risk_7day_prop <- df_final %>%
  filter(
    !is.na(SEX),
    !is.na(CKD),
    !is.na(Alcohol_risk_7day)
  ) %>%
  group_by(SEX, Alcohol_risk_7day) %>%
  summarise(
    proportion_CKD = mean(CKD == "CKD"),
    n = n(),
    .groups = "drop"
  )

ggplot(
  risk_7day_prop,
  aes(
    x = Alcohol_risk_7day,
    y = proportion_CKD,
    fill = SEX
  )
) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "7-day average alcohol risk",
    y = "Proportion with CKD",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    legend.position = "top"
  )

```

Regression using risk levels
```{r}
ckd_filter <- df_final %>%
  filter(
    !is.na(SEX),
    !is.na(CKD),
    !is.na(alcohol_quantity),
    !is.na(Red_wine)
  ) %>%
  mutate(
    CKD = ifelse(CKD == "CKD", 1, 0),
    SEX_factor = factor(
      SEX,
      levels = c("Female", "Male")
    ),
    Alcohol_risk = factor(
      Alcohol_risk,
      levels = c(
        "Low risk",
        "Risky",
        "High risk",
        "No alcohol (past year)",
        "Never alcohol"
      )
    ),
    Alcohol_risk_7day = factor(
      Alcohol_risk_7day,
      levels = c(
        "Low risk",
        "Medium risk",
        "High risk",
        "No alcohol (1 week–<12 months)",
        "No alcohol (≥12 months)",
        "Never alcohol"
      )
    )
  )
```

```{r}
table(ckd_filter$Alcohol_risk_7day, useNA = "ifany") 
```


```{r}
ckd_filter$Beer <- as.integer(
  rowSums(ckd_filter[, c(
    "Low_alcohol_beer",
    "Mid_strength_beer",
    "Full_strength_beer"
  )], na.rm = TRUE) > 0
)

ckd_filter$Wine <- as.integer(
  rowSums(ckd_filter[, c(
    "White_wine",
    "Red_wine",
    "Low_alcohol_wine",
    "Champagne_sparkling_wine",
    "Fortified_wine"
  )], na.rm = TRUE) > 0
)

ckd_filter$Spirits_Liqueurs <- as.integer(
  rowSums(ckd_filter[, c(
    "Spirits_RTD",
    "Liqueurs",
    "Spirits"
  )], na.rm = TRUE) > 0
)

ckd_filter$Other <- as.integer(
  rowSums(ckd_filter[, c(
    "Cider",
    "Cocktails"
  )], na.rm = TRUE) > 0
)
```

```{r}
levels(ckd_filter$Alcohol_risk)
```


```{r}
model_interaction <- glm(CKD ~ alcohol_quantity * SEX_factor + Alcohol_risk * SEX_factor + Beer + Wine + Spirits_Liqueurs + Other, data = ckd_filter, family = binomial)
summary(model_interaction)
```
```{r}
library(emmeans)
```


```{r}
emm <- emmeans(
  model_interaction,
  ~ Alcohol_risk | SEX_factor
)

pairs(
  emm,
  type = "response",
  infer = c(TRUE, TRUE)
)
```

Compare the odds of CKD between different alcohol-risk groups within each sex.  
Among females, Low risk differed significantly from High risk, no significant differences were observed for the other comparisons.  
Among males, Low risk was associated with higher CKD odds than Risky and High risk.  
p: Assuming 2 groups have no differences, the probability of observing this result

OR>1, odds of the previous group is higher -> higher probability of no CKD

```{r}
model_risk_no_int <- glm(
  CKD ~ Alcohol_risk + SEX_factor +
        alcohol_quantity +
        Beer + Wine + Spirits_Liqueurs + Other,
  data = ckd_filter,
  family = binomial
)

model_risk_int <- glm(
  CKD ~ Alcohol_risk * SEX_factor +
        alcohol_quantity +
        Beer + Wine + Spirits_Liqueurs + Other,
  data = ckd_filter,
  family = binomial
)

anova(
  model_risk_no_int,
  model_risk_int,
  test = "Chisq"
)
```
```{r}
model_quantity_no_int <- glm(
  CKD ~ Alcohol_risk + SEX_factor +
        alcohol_quantity +
        Beer + Wine + Spirits_Liqueurs + Other,
  data = ckd_filter,
  family = binomial
)

model_quantity_int <- glm(
  CKD ~ Alcohol_risk +
        alcohol_quantity * SEX_factor +
        Beer + Wine + Spirits_Liqueurs + Other,
  data = ckd_filter,
  family = binomial
)

anova(
  model_quantity_no_int,
  model_quantity_int,
  test = "Chisq"
)
```



